The Talking Heads Experiment

### 2015-04-16 chap4/chap4.tex
hyphenation, orphans, widows and some other layout updated

Categorisation has fascinated philosophers and scientists
since the beginning of thought, not only because it is one
of the most fundamental capabilities of the human mind, 
but also because the subject matter immediately raises 
some intriguing paradoxes and puzzles. First of all, 
we have already seen that there is an enormous 
gap between the symbolic world of objects and categories
and the subsymbolic world of analog sensori-motor signal streams. 
A particular sensory signal is highly context-dependent and
inherently noisy due to the partial
unreliability of sensors and their limited accuracy. 
Sensory processing, transformation, and scaling 
go some way to achieve perceptual constancy
but cognitive processes must clearly make up for
the fleeting erratic nature of reality. 
Here we are immediately confronted with a first paradox. 
Univocal categorisation seems only possible when 
an interpretation of reality is within reach, for example
when there are strong expectations, possibly coming 
from language utterances. But this interpretation
depends itself on categorisation. How can this apparent
causal circularity be broken?

@@

There are basically two philosophical doctrines that 
have tried to address the paradoxes of meaning. 
One doctrine is known as empiricism, the other one as 
rationalism.
Many philosophical texts are available introducing these
philosophical doctrines and their historical roots. 
The problem of the origins of language and meaning
was for example already a highly debated topic among philosophers in the 18th 
century, see for example \cite{Rousseau:1781}.
\newline
The debate between rationalism and empiricism
is still very much alive today and now based
on much more knowledge about what it means for 
something to be innate or what the limits are of
induction. See \cite{Elman:1998}
for the most recent arguments and counterarguments. Compared
to the full richness of human experience, I necessarily 
have had to adopt a very narrow view, focusing only on
basic perceptually grounded categories. More complex
categories related to human relationships, emotions, 
social organisation, or beliefs will not be 
considered and it would be very difficult to do so 
with the methodology used here. See \cite{Varela:1991}
for a broader discussion.
\subsection{The empiricist's stance}

@@

Empiricists argue that categories 
capture what is common or invariant between cases and that
neural networks in the brain
can detect this invariance. They also argue
that these commonalities can be learned
by progressively abstracting away from the details of specific
cases, even if there is a poor stimulus. 
A child sees many examples of red objects and
progressively grasps the abstract concept [RED] 
by retaining what is common to all of them. 
If categories are the result of a general 
inductive learning procedure, the areas of the brain
responsible for categorisation do not have to be
specialised or pre-programmed for recognising 
specific categories. Empiricists therefore believe that 
they can take the 
form of general purpose networks that learn any kind 
of category by making abstraction from the examples 
supplied by the environment. Categories are therefore
not innate but learned.

@@

Empiricists start from the idea that categories 
capture what is common
between objects. Thus [RED] is supposed to capture 
what is common or similar to all 
red objects. Hence in theories of 
(formal) semantics, the meaning of a predicate is 
equated with the set of all things that 
belong to the class it delineates. 
But we can also turn things around. We can 
view a category as a way to capture what is {\itshape different} 
between objects. For example, the distinction between 
[LEFT] and [RIGHT] is based on the horizontal position (HPOS)
of an object with respect to the viewer. 
The distinction is imposed on the scene (as long as 
it is compatible) instead of 
recognised. This may seem a subtle difference, but it
has profound consequences, particularly for acquiring
categories.

Consider the category [LARGE]. What do all large 
objects have in common? At first sight very little. 
Almost any physical object can be called `large' in 
one context or another. 
It is going to be very difficult for a learning agent
to determine some commonality, even if given thousands 
of examples of large objects. In the beginning, the agent
might be confused by commonalities in 
colour or position or shape. 
Only with some luck, will the learning algorithm
start to zoom in on size. 
This is what makes inductive learning so slow and 
why it is criticised, rightfully, by rationalists. 
In fact, searching for commonality hardly makes sense for
many categories. [LARGE] is only meaningful in
opposition to [SMALL], [LEFT] only makes sense
in opposition to [RIGHT]. Often categorisation
is relative to the context (a block which is large when surrounded
by smaller blocks may be categorised as
small when surrounded by larger
blocks) and occasionally it is relative to the objects themselves
(a large mouse is always categorised as
much smaller than a small elephant).

Few people would disagree that [LARGE] and [SMALL] are
distinctions imposed on reality as opposed to intrinsic
properties of classes of objects, but what about colour? Is it
not an example where a category is absolute? We have 
already seen in the previous chapter that this is not 
the case. The colour reflection depends strongly on the 
surface reflection and thus on the light conditions and
light sources in the environment. What should objectively 
appear blue when measuring the wavelength might actually 
appear green and vice-versa. Colour is {\itshape not}
an intrinsic property but is actively mapped onto reality
in a context-sensitive way.\footnote{
As mentioned earlier, see the discussion about colour 
in \cite{Varela:1991}.}
This does not mean of course that categorisation does not make use of sensory
data. On the contrary, without sensory data, categorisation
would be entirely impossible.

Once categories are available, they can be 
used for much more than making distinctions. For example, 
if an agent has the distinction between [LARGE] and [SMALL],
he can use it to group the objects in a scene into two
subsets: those that are large and those that are small. 
So in this case, the categories are used to group 
objects based on a characteristic they have in common, 
namely being large and being small. My main argument here is that 
categories form driven by discrimination tasks,
afterwards they can be used for many other 
semantic processes, including classification.

@@

I will refer to the process capable of making a distinction as
a {\itshape categoriser}. %\cite{categoriser} #FIXME
It operates on the output of 
a sensory channel and decides whether a category or its opponent
is valid. A categoriser keeps track of its success 
by maintaining an internal counter. 
Consider for example the category [LARGE] which 
operates on the output of the AREA-channel. This channel 
contains a scaled value 
between 0.0 and 1.0 for the area of a segment. A distinction 
can therefore be 
made simply by dividing the set of possible values in two
halves: those whose area is between
0.0 and 0.5 and those whose area is between 0.5 and
1.0, giving us two categories, [SMALL] and [LARGE]. When 
the agent needs to categorise an object, he checks in which
region an object's area falls. If it is between 0.0 and 0.5, the
category is [SMALL], if it is greater than 0.5 the category
is [LARGE].

It is clearly possible to refine each 
category $c$ by introducing categorisers that 
further divide the region of possible values of $c$ into 
smaller subregions. For example, the category [SMALL] which 
is applicable if the area is between 0.0 and 0.5, can be refined 
by introducing two more specific categorisers: one responding
to a region between 0.0 and 
0.25 for [VERY-SMALL] and another one for a region
between 0.25 and 0.5 for [MEDIUM-SMALL]. The total set of
distinctions using values of the same sensory channel
can be organised in a discrimination tree (\figref{trees}).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/tree.pdf}}
\caption{\label{trees}A discrimination tree contains
a set of categorisers which categorise by checking whether
a sensory value falls in the region of one category or 
not. The discrimination tree shown operates on
values on the GRAY channel.}
\end{figure}

As mentioned earlier, I will label categories
by using the sensory channel from which a category operates, 
followed by the minimum and maximum values of the region
carved out by the category. For example, 
[AREA 0.0–0.25] carves out the region [0.0,0.25] of the 
area channel. When it must be emphasised that 
a category belongs to a particular agent, for 
example {\bfshape a1}, I will write [AREA 0.0–0.25]$_{\mathbf a1}$.

Conjunctive combinations of categories also have
dedicated categorisers which are linked to 
the categorisers of their components (see 
\figref{disnet}). A conjunctive combination 
often yields a more efficient way to pick out 
the topic compared to a single, possibly very fine-grained
distinction. For example, it might be that 
[AREA 0.0–0.25] {\itshape and} [GRAY 0.5–1.0]
together are distinctive but none of the two on their
own is. Other logical combinations are equally of interest
but I will restrict my attention to conjunctive 
combinations. Conjunctive combinations of categories 
will be written between curly brackets (\{,\}) as in 
\{[AREA 0.0–0.25] [GRAY 0.5–1.0]\}.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/game5.pdf}}
\caption{\label{geom}A computer generated scene 
from the GEOM world.}
\end{figure}

I will now develop some concrete examples, which imply
that I make choices for the kinds of sensory channels
and scenes that the agents use. 
I will first use scenes from the GEOM world as in 
\figref{geom} and later real world scenes captured
with a Talking Heads camera.

Consider the scene in \figref{geom}. Assuming that the agent
already has a well-developed set of categories, he could use 
the category [GRAY-0.75,1.0] (very dark) to distinguish 
shape 3 from the others. On the other hand, if shape 2 is 
chosen as topic, the grayscale will not be enough because
shape 2 has the same grayscale as shape 1. Maybe a 
combination of categories can be chosen, like [VPOS 0.0–05]
(lower), and [HPOS 0.0–0.5] (left). Indeed shape 2
is lower in the scene, as opposed to
shape 1 and shape 3, and it is more to the left compared 
to shape 0 and 2.

@@

Imagine that for each object in
the scene and for each sensory channel, there is a ball
containing the value for that object on that channel. It
is introduced in the top categoriser of the 
discrimination tree associated with that channel. 
For example, suppose that there are three
objects in the scene: $O1$, $O2$, and $O3$, 
with gray-scale values 0.6, 0.4, and 0.9 respectively. 
We can therefore imagine three balls labeled with these datavalues
which are input to the top categoriser of the 
GRAY discrimination tree (\figref{balls}). 
A categoriser divides the balls in two bins, 
those that fall in the range of 
one category and those that fall into the range of the other
category. In this case, the left bin contains \{$O2$\} 
(category [GRAY 0.0–0.5]) 
and the right bin \{$O1$, $O3$\} (category [GRAY 0.5–1.0]).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/balls.pdf}}
\caption{\label{balls}Balls containing
the data value for the different objects
$O1$, $O2$, $O3$ in the scene. $O1$ is the topic.
$O1$ has the value 0.6 on the GRAY-channel, $O2$ has
0.4 and $O3$ 0.9.} 
\end{figure}

A distinctive category is found when the ball of the 
topic is the only one left in one of the bins. 
If the topic is $O1$, then this is not yet the 
case, because it is together with $O3$ in the 
bin of [GRAY 0.5–1.0]. However when the next categoriser
is exercised, it splits the set \{$O1$, $O3$\} 
into two subsets: \{$O1$\} for [GRAY 0.5–0.75]
and \{$O3$\} for [GRAY-0.75,1.0]. 
[GRAY 0.5–0.75] is a discriminating category because only 
$O1$ is left in its bin.

@@

There are obviously many variations on categorisation
that could be imagined. For example, a categoriser
could make use of focal points instead of regions. A focal 
point is a single significant data value of a sensory channel.
The categoriser then has to compute the distance 
between the value for a segment and the focal point 
of each possible category. The category 
whose focal point is closest to the sensory value
of a segment applies. This implements a prototype-like
approach to categorisation which has been argued to 
be more realistic with respect to human categorisation.\footnote{
See \cite{Varela:1991} and \cite{Taylor:1989}.}
For example, humans typically label light
at 482 nanometres as the most typical blue, so that a given object
reflecting light at or near this point is categorised
as [BLUE]. Categories based on focal points are 
interesting and have clear advantages 
but I will stick nevertheless in the first instance 
to binary discrimination trees operating on single sensory 
channels to simplify the explanations
and to analyse better what is going on.

Still another way to categorise reality is by imposing 
an order on the segments based on their values for 
a particular sensory channel. For example, we can 
order the segments based on the HPOS channel (i.e. 
from left to right) and then introduce relational categories
like `left of one segment in the series', or `the left-most object'. 
Similarly we can order the segments based on the 
HEIGHT channel (i.e. in terms of their size) and then 
have categories that select the smallest (i.e. the 
first segment in this ordering), or those greater than 
some other one.

@@

Let us now look at some example 
games for an agent {\bfshape a1} taken from simulations
using computer generated
scenes from the GEOM world and showing the internal
structures generated as well as the reports from the commentator. 
At first I will not take saliency nor 
context-scaling into account. The first game
(Game 8) fails. It takes place near the very beginning 
when {\bfshape a1} has practically 
no repertoire of distinctions yet. The scene and the 
discrimination trees available so far are shown in Figure 
\ref{scene2}. The object labeled 0 is the
topic. Only two channels have top level
categorisers: HEIGHT and WIDTH. 
The data on these channels for the scene in Figure
\ref{scene2} are shown in \tabref{tab:t-game8}.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HEIGHT} & {\itshape WIDTH} \\ \midrule
0 (square) & 0.413 & 0.317  \\ 
1 (circle) & 0.410 & 0.410 \\ 
2 (square) & 0.163 & 0.163 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game8}Sensory data for the scene in \figref{scene2}.}
\end{center}
\end{table}

HEIGHT and WIDTH have been scaled with respect to the 
minimum and maximum height and width of a 
figure (sensor-scaling) but no context-scaling has 
been performed.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/scene22.pdf}}
\caption{\label{scene3}Top: The scene used in Game 22. 
Bottom: The discrimination trees available to the
agent.}
\end{figure}
The next example shows a game (Game 22) based on the scene 
in \figref{scene3}. The topic is the triangle, shape 0. 
The discrimination trees for 
HEIGHT and WIDTH have already more than one level and 
a discrimination tree for HPOS has developed.
The relevant sensor-scaled data for these three sensory
channels is shown in \tabref{tab:t-game22}.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HPOS} & {\itshape WIDTH} & {\itshape HEIGHT} \\ \midrule
0 (triangle) & 0.167 & 0.437 & 0.573  \\ 
1 (rectangle) & 0.789 & 0.563 & 0.287 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22}Sensory data for the scene shown in \figref{scene3}.}
\end{center}
\end{table}

@@

Game 22 shows at once why context-scaling 
and saliency is important. When we inspect the scene in 
\figref{scene3}, we do not quite see so clearly 
that the triangle is less wide than the square, so why
is [WIDTH 0.0–0.5] 
nevertheless considered? Examination of the data shows
that the WIDTH values, 0.437 for the triangle
and 0.563 for the square, are very close 
to each other, but just by luck fall within 
the two regions carved out by the WIDTH discrimination
tree. On the other hand, the values on the HPOS channel 
are much further apart and so they are preferred.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape HPOS} & {\itshape WIDTH} & {\itshape HEIGHT} \\ \midrule
 0.622 & 0.125 & 0.286  \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22-sal}Sensory data for the scene in Game 22.}
\end{center}
\end{table}
From this table we see immediately that HPOS is the 
most salient channel and should be preferred by far, followed
by HEIGHT and then WIDTH. Thus we can expect the
agent to choose HPOS based on saliency. When the saliency threshold
is set to a reasonably high value, the other channels
would not even be considered, they would not pass the 
sieve of the perceptual layer.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HPOS} & {\itshape WIDTH} & {\itshape HEIGHT} \\ \midrule
0 (triangle) & 0.0 & 0.0 & 1.0  \\ 
1 (rectangle) & 1.0 & 1.0 & 0.0 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22scaled}Sensory data for the scene shown in \figref{scene3}.}
\end{center}
\end{table}
So the best thing to do (and this is what the Talking Heads 
effectively do) is to first perform sensor-scaling, then 
compute saliency to determine which channel should be 
preferred, then perform context-scaling, to get clearly 
distinguished sensory values, and then do categorisation. 
Note that context-scaling has the same effect as using 
prototype-based categorisation because the actual values 
are pulled towards extremes, and thus perceived as 
prototypes. Context-scaling is not always desirable. For example, 
in the case of colour categorisation the actual channel
data should be maintained because here categorisation 
takes place on the basis of actual values.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/game24.pdf}}
\caption{\label{scene4}The scene used in Game 24.}
\end{figure}
The relevant data after sensor-scaling for the 
two sensory channels involved are shown in \tabref{tab:t-game24}.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HEIGHT} & {\itshape WIDTH} \\ \midrule
0 (triangle) & 0.170 & 0.513  \\ 
1 (triangle) & 0.653 & 0.570 \\  
2 (square) & 0.213 & 0.213 \\  
3 (rectangle) & 0.613 & 0.310 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game24}Sensory data for the scene in \figref{scene4}.}
\end{center}
\end{table}

[HEIGHT 0.0–0.5] is valid for triangle-0 and square-2 but
filters out the other segments. 
[WIDTH 0.5–1.0] is valid for triangle-0 and triangle-1 and 
filters out the others.  
The conjunctive combination of these two categories only 
retains triangle-0 and is therefore the one that is chosen.

>>>>

Categorisation has fascinated philosophers and scientists
since the onset \todo{beginning was replaced with onset here} of thought, not only because it is one
of the most fundamental capabilities of the human mind, 
but also because the subject matter immediately raises 
some intriguing paradoxes and puzzles. First of all, 
we have already seen that there is an enormous 
gap between the symbolic world of objects and categories
and the subsymbolic world of analog sensori-motor signal streams. 
A particular sensory signal is highly context-dependent and
inherently noisy due to the partial
unreliability of sensors and their limited accuracy. 
Sensory processing, transformation, and scaling 
go some way to achieve perceptual constancy
but cognitive processes must clearly make up for
the fleeting erratic nature of reality. 
Here we are immediately confronted with a first paradox. 
Univocal categorisation seems only possible when 
an interpretation of reality is within reach, for example
when there are strong expectations, possibly coming 
from language utterances. But this interpretation
depends itself on categorisation. How can this apparent
causal circularity be broken?

@@

There are basically two philosophical doctrines that 
have tried to address the paradoxes of meaning. 
One doctrine is known as empiricism, the other one as 
rationalism.
Many philosophical texts are available introducing these
philosophical doctrines and their historical roots. 
The problem of the origins of language and meaning
was for example already a highly debated topic among philosophers in the 18th 
century, see for example \cite{Rousseau:1781}.
\newline \enlargethispage{1\baselineskip}
The debate between rationalism and empiricism
is still very much alive today and now based
on much more knowledge about what it means for 
something to be innate or what the limits are of
induction. See \cite{Elman:1998}
for the most recent arguments and counterarguments. Compared
to the full richness of human experience, I necessarily 
have had to adopt a very narrow view, focusing only on
basic perceptually grounded categories. More complex
categories related to human relationships, emotions, 
social organisation, or beliefs will not be 
considered and it would be very difficult to do so 
with the methodology used here. See \cite{Varela:1991}
for a broader discussion.
\subsection{The empiricist's stance}

@@

Empiricists argue that categories 
capture what is common or invariant between cases and that
neural networks in the brain
can detect this invariance. They also argue
that these commonalities can be learned
by progressively abstracting away from the details of specific
cases, even if there is a poor stimulus. 
A child sees many examples of red objects and
progressively grasps the abstract concept [\textsc{red}] 
by retaining what is common to all of them. 
If categories are the result of a general 
inductive learning procedure, the areas of the brain
responsible for categorisation do not have to be
specialised or pre-programmed for recognising 
specific categories. Empiricists therefore believe that 
they can take the 
form of general purpose networks that learn any kind 
of category by making abstraction from the examples 
supplied by the environment. Categories are therefore
not innate but learned.

@@

Empiricists start from the idea that categories 
capture what is common
between objects. Thus [\textsc{red}] is supposed to capture 
what is common or similar to all 
red objects. Hence in theories of 
(formal) semantics, the meaning of a predicate is 
equated with the set of all things that 
belong to the class it delineates. 
But we can also turn things around. We can 
view a category as a way to capture what is {\itshape different} 
between objects. For example, the distinction between 
[\textsc{left}] and [\textsc{right}] is based on the horizontal position (\textsc{hpos})
of an object with respect to the viewer. 
The distinction is imposed on the scene (as long as 
it is compatible) instead of 
recognised. This may seem a subtle difference, but it
has profound consequences, particularly for acquiring
categories.

Consider the category [\textsc{large}]. What do all large 
objects have in common? At first sight very little. 
Almost any physical object can be called `large' in 
one context or another. 
It is going to be very difficult for a learning agent
to determine some commonality, even if given thousands 
of examples of large objects. In the beginning, the agent
might be confused by commonalities in 
colour or position or shape. 
Only with some luck, will the learning algorithm
start to zoom in on size. 
This is what makes inductive learning so slow and 
why it is criticised, rightfully, by rationalists. 
In fact, searching for commonality hardly makes sense for
many categories. [\textsc{large}] is only meaningful in
opposition to [\textsc{small}], [\textsc{left}] only makes sense
in opposition to [\textsc{right}]. Often categorisation
is relative to the context (a block which is large when surrounded
by smaller blocks may be categorised as
small when surrounded by larger
blocks) and occasionally it is relative to the objects themselves
(a large mouse is always categorised as
much smaller than a small elephant).

Few people would disagree that [\textsc{large}] and [\textsc{small}] are
distinctions imposed on reality as opposed to intrinsic
properties of classes of objects, but what about colour? Is it
not an example where a category is absolute? We have 
already seen in the previous chapter that this is not 
the case. The colour reflection depends strongly on the 
surface reflection and thus on the light conditions and
light sources in the environment. What should objectively 
appear blue when measuring the wavelength might actually 
appear green and vice-versa. Colour is {\itshape not}
an intrinsic property but is actively mapped onto reality
in a context-sensitive way.\footnote{
As mentioned earlier, see the discussion about colour 
in \cite{Varela:1991}.}
This does not mean of course that categorisation does not make use of sensory
data. On the contrary, without sensory data, categorisation
would be entirely impossible.

Once categories are available, they can be 
used for much more than making distinctions. For example, 
if an agent has the distinction between [\textsc{large}] and [\textsc{small}],
he can use it to group the objects in a scene into two
subsets: those that are large and those that are small. 
So in this case, the categories are used to group 
objects based on a characteristic they have in common, 
namely being large and being small. My main argument here is that 
categories form driven by discrimination tasks,
afterwards they can be used for many other 
semantic processes, including classification.

@@

I will refer to the process capable of making a distinction as
a {\itshape categoriser}. %\cite{categoriser} #FIXME
It operates on the output of 
a sensory channel and decides whether a category or its opponent
is valid. A categoriser keeps track of its success 
by maintaining an internal counter. 
Consider for example the category [\textsc{large}] which 
operates on the output of the \textsc{area}-channel. This channel 
contains a scaled value 
between 0.0 and 1.0 for the area of a segment. A distinction 
can therefore be 
made simply by dividing the set of possible values in two
halves: those whose area is between
0.0 and 0.5 and those whose area is between 0.5 and
1.0, giving us two categories, [\textsc{small}] and [\textsc{large}]. When 
the agent needs to categorise an object, he checks in which
region an object's area falls. If it is between 0.0 and 0.5, the
category is [\textsc{small}], if it is greater than 0.5 the category
is [\textsc{large}].

It is clearly possible to refine each 
category $c$ by introducing categorisers that 
further divide the region of possible values of $c$ into 
smaller subregions. For example, the category [\textsc{small}] which 
is applicable if the area is between 0.0 and 0.5, can be refined 
by introducing two more specific categorisers: one responding
to a region between 0.0 and 
0.25 for [\textsc{very}-\textsc{small}] and another one for a region
between 0.25 and 0.5 for [\textsc{medium}-\textsc{small}]. The total set of
distinctions using values of the same sensory channel
can be organised in a discrimination tree (\figref{trees}).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/tree.pdf}}
\caption{\label{trees}A discrimination tree contains
a set of categorisers which categorise by checking whether
a sensory value falls in the region of one category or 
not. The discrimination tree shown operates on
values on the \textsc{gray} channel.}
\end{figure}

As mentioned earlier, I will label categories
by using the sensory channel from which a category operates, 
followed by the minimum and maximum values of the region
carved out by the category. For example, 
[\textsc{area} 0.0–0.25] carves out the region [0.0,0.25] of the 
area channel. When it must be emphasised that 
a category belongs to a particular agent, for 
example {\bfshape a1}, I will write [\textsc{area} 0.0–0.25]$_{\mathbf a1}$.

Conjunctive combinations of categories also have
dedicated categorisers which are linked to 
the categorisers of their components (see 
\figref{disnet}). A conjunctive combination 
often yields a more efficient way to pick out 
the topic compared to a single, possibly very fine-grained
distinction. For example, it might be that 
[\textsc{area} 0.0–0.25] {\itshape and} [\textsc{gray} 0.5–1.0]
together are distinctive but none of the two on their
own is. Other logical combinations are equally of interest
but I will restrict my attention to conjunctive 
combinations. Conjunctive combinations of categories 
will be written between curly brackets (\{,\}) as in 
\{[\textsc{area} 0.0–0.25] [\textsc{gray} 0.5–1.0]\}.

@@

\begin{figure}[b]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/game5.pdf}}
\caption{\label{geom}A computer generated scene 
from the \textsc{geom} world.}
\end{figure}

I will now develop some concrete examples, which imply
that I make choices for the kinds of sensory channels
and scenes that the agents use. 
I will first use scenes from the \textsc{geom} world as in 
\figref{geom} and later real world scenes captured
with a Talking Heads camera.

Consider the scene in \figref{geom}. Assuming that the agent
already has a well-developed set of categories, he could use 
the category [\textsc{gray}-0.75,1.0] (very dark) to distinguish 
shape 3 from the others. On the other hand, if shape 2 is 
chosen as topic, the grayscale will not be enough because
shape 2 has the same grayscale as shape 1. Maybe a 
combination of categories can be chosen, like [\textsc{vpos} 0.0–05]
(lower), and [\textsc{hpos} 0.0–0.5] (left). Indeed shape 2
is lower in the scene, as opposed to
shape 1 and shape 3, and it is more to the left compared 
to shape 0 and 2.

@@

Imagine that for each object in
the scene and for each sensory channel, there is a ball
containing the value for that object on that channel. It
is introduced in the top categoriser of the 
discrimination tree associated with that channel. 
For example, suppose that there are three
objects in the scene: \emph{O1}, \emph{O2}, and \emph{O3}, 
with gray-scale values 0.6, 0.4, and 0.9 respectively. 
We can therefore imagine three balls labeled with these datavalues
which are input to the top categoriser of the 
\textsc{gray} discrimination tree (\figref{balls}). 
A categoriser divides the balls in two bins, 
those that fall in the range of 
one category and those that fall into the range of the other
category. In this case, the left bin contains \{\emph{O2}\} 
(category [\textsc{gray} 0.0–0.5]) 
and the right bin \{\emph{O1}, \emph{O3}\} (category [\textsc{gray} 0.5–1.0]).

\begin{figure}[b]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/balls.pdf}}
\caption{\label{balls}Balls containing
the data value for the different objects
\emph{O1}, \emph{O2}, \emph{O3} in the scene. \emph{O1} is the topic.
\emph{O1} has the value 0.6 on the \textsc{gray}-channel, \emph{O2} has
0.4 and \emph{O3} 0.9.} 
\end{figure}

A distinctive category is found when the ball of the 
topic is the only one left in one of the bins. 
If the topic is \emph{O1}, then this is not yet the 
case, because it is together with \emph{O3} in the 
bin of [\textsc{gray} 0.5–1.0]. However when the next categoriser
is exercised, it splits the set \{\emph{O1}, \emph{O3}\} 
into two subsets: \{\emph{O1}\} for [\textsc{gray} 0.5–0.75]
and \{\emph{O3}\} for [\textsc{gray}-0.75,1.0]. 
[\textsc{gray} 0.5–0.75] is a discriminating category because only 
\emph{O1} is left in its bin.

@@

There are obviously many variations on categorisation
that could be imagined. For example, a categoriser
could make use of focal points instead of regions. A focal 
point is a single significant data value of a sensory channel.
The categoriser then has to compute the distance 
between the value for a segment and the focal point 
of each possible category. The category 
whose focal point is closest to the sensory value
of a segment applies. This implements a prototype-like
approach to categorisation which has been argued to 
be more realistic with respect to human categorisation.\footnote{
See \cite{Varela:1991} and \cite{Taylor:1989}.}
For example, humans typically label light
at 482 nanometres as the most typical blue, so that a given object
reflecting light at or near this point is categorised
as [\textsc{blue}]. Categories based on focal points are 
interesting and have clear advantages 
but I will stick nevertheless in the first instance 
to binary discrimination trees operating on single sensory 
channels to simplify the explanations
and to analyse better what is going on.

Still another way to categorise reality is by imposing 
an order on the segments based on their values for 
a particular sensory channel. For example, we can 
order the segments based on the \textsc{hpos} channel (i.e. 
from left to right) and then introduce relational categories
like `left of one segment in the series', or `the left-most object'. 
Similarly we can order the segments based on the 
\textsc{height} channel (i.e. in terms of their size) and then 
have categories that select the smallest (i.e. the 
first segment in this ordering), or those greater than 
some other one.

@@

Let us now look at some example 
games for an agent {\bfshape a1} taken from simulations
using computer generated
scenes from the \textsc{geom} world and showing the internal
structures generated as well as the reports from the commentator. 
At first I will not take saliency nor 
context-scaling into account. The first game
(Game 8) fails. It takes place near the very beginning 
when {\bfshape a1} has practically 
no repertoire of distinctions yet. The scene and the 
discrimination trees available so far are shown in Figure 
\ref{scene2}. The object labeled 0 is the
topic. Only two channels have top level
categorisers: \textsc{height} and \textsc{width}. 
The data on these channels for the scene in Figure
\ref{scene2} are shown in \tabref{tab:t-game8}.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape Object} & {\itshape \textsc{height}} & {\itshape \textsc{width}} \\ \midrule
0 (square) & 0.413 & 0.317  \\ 
1 (circle) & 0.410 & 0.410 \\ 
2 (square) & 0.163 & 0.163 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game8}Sensory data for the scene in \figref{scene2}.}
\end{center}
\end{table}

\textsc{height} and \textsc{width} have been scaled with respect to the 
minimum and maximum height and width of a 
figure (sensor-scaling) but no context-scaling has 
been performed.

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape \textsc{hpos}} & {\itshape \textsc{width}} & {\itshape \textsc{height}} \\ \midrule
0 (triangle) & 0.167 & 0.437 & 0.573  \\ 
1 (rectangle) & 0.789 & 0.563 & 0.287 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22}Sensory data for the scene shown in \figref{scene3}.}
\end{center}
\end{table}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/scene22.pdf}}
\caption{\label{scene3}Top: The scene used in Game 22. 
Bottom: The discrimination trees available to the
agent.}
\end{figure}
\enlargethispage{1\baselineskip}
The next example shows a game (Game 22) based on the scene 
in \figref{scene3}. The topic is the triangle, shape 0. 
The discrimination trees for 
\textsc{height} and \textsc{width} have already more than one level and 
a discrimination tree for \textsc{hpos} has developed.
The relevant sensor-scaled data for these three sensory
channels is shown in \tabref{tab:t-game22}.

@@

Game 22 shows at once why context-scaling 
and saliency is important. When we inspect the scene in 
\figref{scene3}, we do not quite see so clearly 
that the triangle is less wide than the square, so why
is [\textsc{width} 0.0–0.5] 
nevertheless considered? Examination of the data shows
that the \textsc{width} values, 0.437 for the triangle
and 0.563 for the square, are very close 
to each other, but just by luck fall within 
the two regions carved out by the \textsc{width} discrimination
tree. On the other hand, the values on the \textsc{hpos} channel 
are much further apart and so they are preferred.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape \textsc{hpos}} & {\itshape \textsc{width}} & {\itshape \textsc{height}} \\ \midrule
 0.622 & 0.125 & 0.286  \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22-sal}Sensory data for the scene in Game 22.}
\end{center}
\end{table}
From this table we see immediately that \textsc{hpos} is the 
most salient channel and should be preferred by far, followed
by \textsc{height} and then \textsc{width}. Thus we can expect the
agent to choose \textsc{hpos} based on saliency. When the saliency threshold
is set to a reasonably high value, the other channels
would not even be considered, they would not pass the 
sieve of the perceptual layer.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape \textsc{hpos}} & {\itshape \textsc{width}} & {\itshape \textsc{height}} \\ \midrule
0 (triangle) & 0.0 & 0.0 & 1.0  \\ 
1 (rectangle) & 1.0 & 1.0 & 0.0 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22scaled}Sensory data for the scene shown in \figref{scene3}.}
\end{center}
\end{table}
So the best thing to do (and this is what the Talking Heads 
effectively do) is to first perform sensor-scaling, then 
compute saliency to determine which channel should be 
preferred, then perform context-scaling, to get clearly 
distinguished sensory values, and then do categorisation. 
Note that context-scaling has the same effect as using 
prototype-based categorisation because the actual values 
are pulled towards extremes, and thus perceived as 
prototypes. Context-scaling is not always desirable. For example, 
in the case of colour categorisation the actual channel
data should be maintained because here categorisation 
takes place on the basis of actual values.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/game24.pdf}}
\caption{\label{scene4}The scene used in Game 24.}
\end{figure}
The relevant data after sensor-scaling for the 
two sensory channels involved is shown in \tabref{tab:t-game24}.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape \textsc{height}} & {\itshape \textsc{width}} \\ \midrule
0 (triangle) & 0.170 & 0.513  \\ 
1 (triangle) & 0.653 & 0.570 \\  
2 (square) & 0.213 & 0.213 \\  
3 (rectangle) & 0.613 & 0.310 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game24}Sensory data for the scene in \figref{scene4}.}
\end{center}
\end{table}

[\textsc{height} 0.0–0.5] is valid for triangle-0 and square-2 but
filters out the other segments. 
[\textsc{width} 0.5–1.0] is valid for triangle-0 and triangle-1 and 
filters out the others.  
The conjunctive combination of these two categories only 
retains triangle-0 and is therefore the one that is chosen.
